Supplementary Materials

### 2023-07-21 manuscript/manuscript.tex
ms updates from the frankfurt airport

Prior work to model the reduction in linked diversity due to deleterious
mutations has largely relied on the classic Background Selection (BGS) model
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1995-xc}.
While this model has been successful in explaining many patterns of diversity,
some its simplifying assumptions may distort inferences about the negative
selection process. First, since fixation probabilities ultimately depend on the
product of the deleterious selection coefficient ($s$) and population size
($N$), the efficacy of negative selection should depend on demography.
Unfortunately, accommodating demography into theoretic negative selection
models remains an open, difficult problem \parencite{Zeng2013-ep,Johri2020-oj}.
Second, building off classic models of mutation-selection balance
\parencite{Crow1970-wm,Kimura1966-bk}, the BGS model assumes that new mutations
are sufficiently deleterious that they are invariably driven to loss. Under
this assumption, the effect of selection is well-approximated by simply
rescaling the neutral coalescent by a reduction factor known as $B =
\nicefrac{N}{N_e}$ \parencite{Charlesworth2013-kl}. However, this simple
rescaling approach is not appropriate across parts of parameter space that are
relevant to natural populations \parencite{McVean2000-bt,Good2014-yz}. In
particular, the BGS model cannot accommodate the possibility of fixation for
mutations with $2Ns \ll 1$, leading diversity to be incorrectly predicted to
fall to zero as the strength of selection diminishes. This weak selection
problem is related to the wider problem of modeling the substitution rate of
deleterious mutation (i.e. the ``ratchet" rate) and incorporating the impact of
interference from other neighboring selected sites.

@@

Similar to previous work, we aim to characterize the selection processes
through the signal left on nucleotide diversity at linked sites. Although
nucleotide diversity is a less-informative statistic than the full site
frequency spectrum, it is much easier to derive theoretic models of how linked
selection alters expected pairwise coalescence times than to describe the
entire genealogy. Still, inferences drawn from the linked selection signal
about selective processes are only as accurate as the underlying theoretic
model. Our work is motivated by a well-known limitation of classic background
selection theory: it only accurately predicts levels of linked diversity when
new mutations are strongly deleterious
\parencite{Charlesworth1993-gb,McVean2000-bt,Good2013-lp,Gordo2002-dr}.
Unfortunately, modifying background selection theory so that it accommodates
weak selection has been challenging problem
\parencite{Good2014-yz,Haigh1978-gt,Higgs1995-xc}. Here, we develop intuition
for why the classic BGS model works well for strong selection but not weak, and
how formulating the equilibrium between new deleterious mutations and negative
selection in terms of the additive fitness variance can alleviate these
problems.

The theoretic basis of our method models the effects of weak draft on diversity
in terms of polygenic additive fitness variance. This fitness variance could
occur due to any weakly selected

Throughout, we imagine a negative selection process where a continual influx of
deleterious mutations occur at a rate of $\mu$ per basepair per generation in a
conserved region of $L$ basepairs, such that the region-wide per generation
diploid mutation rate is $U = 2 \mu L$. Each mutation imposes a selective cost
of $s$ in heterozygotes and $2s$ in homozygotes, and fitness effects are
multiplicative across sites. When selection against a new mutation is strong,
the stochastic perturbations it experiences due to random genetic drift are
weak relative to the frequency changes caused by selection. In this case, every
deleterious mutation is destined to eventual extinction, and only the fraction
of lineages without any mutations contribute to the long-run ancestry of future
populations. The classic background selection approximates this process by
rescaling the neutral coalescent with an effective population size of $N_0 = N
\exp(-\nicefrac{U}{s})$
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1994-oh}.
This approximation works well for two reasons (though see
\cite{Cvijovic2018-vd,Walczak2012-fi,Nicolaisen2012-vs}). First, the number of
mutations per haplotype has has a stationary Poisson distribution with an
average equal to the deterministic mutation-selection equilibrium
\parencite{Haldane1927-ga}. Second, the time it takes for a present-day lineage
to trace their ancestry back to a mutation-free ancestor (the ``delay phase")
is negligible compared to the coalescence timescale of $N_0$ among
mutation-free ancestors (\cite{Durrett2008-ql}, p. 213; \cite{Good2014-yz}).

By contrast, when selection is weak, the frequency trajectories of both neutral
and selected sites are strongly impacted by stochastic perturbations from two
sources. First, genetic drift, which is due to Mendelian segregation and
non-heritable fitness differences; the magnitude of its perturbations are set
by the \emph{drift-effective} population size, $N_e$. Second, genetic draft
occurs when an allele becomes randomly associated with a fitness background
that perturbs its trajectory across generations
\parencite{Neher2013-dz,Gillespie2000-mh,Gillespie2001-mv}. Like genetic drift,
the perturbations due to draft are directionless but overall increase the
variance in allele frequency change. Unlike drift, the impact of the stochastic
perturbations due to draft are autocorrelated and build up over generations
until the fitness background recombines off
\parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. The stochastic
perturbations due to selection also impact selected sites, and alter their
dynamics, creating Hill--Robertson interference \parencite{Hill1966-kd}. Strong
draft can generate multiple-merger coalescences that create discontinuities in
allele frequency trajectories, making diffusion approximations ill-suited
\parencite{Gillespie2000-mh,Der2011-it,Neher2013-dz}. However, when selection
is weak, draft is well-approximated as an increase in the rate of drift (or
equivalently, a reduction in effective population size).

The stochastic perturbations from genetic drift and draft impact both
genealogies and selection dynamics in complex ways, which has made finding
analytic theory predicting their joint effects difficult. When drift acts
alone, the relevant scale of perturbations is determined by the
\emph{drift-effective population size} $N_e$
\parencite{Ohta1971-gq,Ohta1992-yi}. When the random perturbations due to drift
are stronger than the changes due to selection ($2N_es \le 1$), weakly
deleterious mutations can drift up to intermediate frequencies before their
eventual loss or fixation. The assumptions of classic BGS suddenly breakdown in
two ways. First, the ``delay phase" is no longer negligible and genealogies are
no longer well-approximated by a rescaled neutral coalescence process
\parencite{Przeworski1999-mb,OFallon2010-my,Higgs1995-xc}. Second, the
distribution of the number of deleterious mutations (and its corresponding
fitness distribution) is no longer a stationary Poisson distribution, instead
becoming a traveling wave \parencite{Rouzine2008-qz,Good2013-lp,Gessler1995-hz}
towards reduced population fitness (since we ignore beneficial mutations). When
this occurs, individuals in the least-loaded class are stochastically lost,
fixing those mutations with a click of ``Muller's ratchet"
\parencite{Muller1964-ki,Charlesworth1997-qn}. Unfortunately, determining the
rate of Muller's ratchet is another difficult, open problem
\parencite{Haigh1978-gt,Gordo2002-dr,Gessler1995-hz} related to Hill--Robertson
interference \parencite{Felsenstein1974-xm}.

Our work extends theory from \textcite{Santiago2016-mu}, which suggests both
strong and weak negative selection can be be accommodated with a quantitative
genetic model of linked selection. These linked selection models stem from
\textcite{Robertson1961-ho}, which in essence describes how heritable polygenic
fitness variation generates weak genetic draft. At the individual level
Robertson considered, selection generates autocorrelation as offspring from
large families tend to beget many descendents themselves (and likewise with
small families) when fitness is heritable. This same autocorrelation occurs at
the genomic level with linkage \parencite{Santiago1998-bs,Barton2000-zg}, as
the perturbations to a neutral allele's trajectory from its particular fitness
background tend to occur in the same direction across generations until the
background recombines off. These models quantify the total impact of the
autocorrelation generated by weak draft in terms of what we think of as a
\emph{draft-effective} population size $N_d$. The key insight is that in the
long-run, steady, weak draft created by the presence of additive genetic
fitness variance $V_A > 0$ acts like an extra variance in offspring number
would under pure drift \parencite{Wright1938-tv}. However, due to
autocorrelation, the cumulative effect of this fitness variance is inflated by
a factor of $Q^2$. Intuitively, the product $V_A Q^2$ represents the expected
total variance in reproductive success a neutral mutation experiences over its
lifetime in a system with weak selection draft.

We define the draft-effective population size $N_d$ by including the total
additional variance created by heritable fitness in Wright's
(\citeyear{Wright1938-tv}) equation for effective population size,

@@

Equation \eqref{eq:main_Ne} is general, since different modes of selection and
linkage can be accommodated by different expressions for $V_A$ and the
inflation factor $Q^2$ \parencite{Santiago1995-hx,Santiago1998-bs}. When
fitness variation has a multiplicative polygenic basis, as is often assumed for
genome-wide negative selection processes, the draft asymptotic effective
population size experienced by an arbitrary neutral site under the influence of
all linked regions is,

@@

The remaining pieces are expressions for the equilibrium additive fitness
variance $V_A$ and the decay rate in associated fitness $1-\kappa$. At this
point, we diverge from Santiago and Caballero (\citeyear{Santiago1998-bs},
\citeyear{Santiago2016-mu}) to note that the additive fitness variation is the
sum of additive \emph{genic} fitness variance $V_a$ (i.e. due to allele
frequencies) and the fitness covariance due to linkage disequilibria between
selected sites ($\delta_{LD}$), $V_A = V_a + \delta_{LD}$. Considering only the
genic variation, at equilibrium $\Delta V_{a,i} = 0$; thus, the loss in genic
fitness each generation due to selective dynamics $-\kappa V_{A,i}$ must be
equal to the increase in fitness variation due to new mutations each generation
($V_m$) and changes in the LD term \parencite{Bulmer1971-ae}. Note that the
loss in genic fitness due to drift is not time-invariant due to the build up of
autocorrelation, and considered elsewhere in the derivation (see Supplementary
Materials Equation \ref{eq:vardecay}). Then, at equilibrium $\kappa_i =
\nicefrac{V_{m,i}}{V_{a,i}}$ (see Supplementary Materials Equation \ref{eq:Z}).

Under any selection model, the additive genic fitness variance created by a new
mutation (at frequency $x=\nicefrac{1}{2N}$) is $2s^2 x(1-x) \approx
\nicefrac{s^2}{N}$. For the entire population of $2N$ chromosomes, the fresh
variance from mutation each generation in segment $i$ is $V_{m,i} \approx
U_is^2$ where $U_i = 2\mu L_i$ is the diploid mutation rate per generation
within the segment. Under mutation-selection balance assumed by BGS, an
$L_i$-basepair segment has genic variance $V_{a,i}^{BGS} \approx U_i s$ (see
Supplementary Materials Equation \ref{supp-eq:va_bgs}) and thus $\kappa_i^{BGS}
= s$. Substituting $V_{a,i}^{BGS}$ and $\kappa_i^{BGS}$ in Equation
\eqref{eq:polygenic_Ne} and simplifying, we have

@@

When selection is weak, however, the additive fitness variance under the
classic BGS model is no longer accurate. \textcite{Santiago2016-mu} suggested
that this is due to the loss of fitness variation that occurs when a
segregating site fixes and its heterozygosity becomes zero. If we let $R$
represent the fixation rate (i.e. ratchet rate) in the region per generation,
each fixation removes the equivalent amount of equilibrium variation put in by
mutation. Thus, the steady-state genic variance under mutation and negative
selection is (omitting the segment index $i$ for clarity),

>>>>

Prior work to model the reduction in linked diversity due to deleterious
mutations has largely relied on the classic Background Selection (BGS) model
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1995-xc}.
While this model has been successful in explaining many patterns of diversity,
some its simplifying assumptions may distort inferences about the negative
selection process. First, since fixation probabilities ultimately depend on the
product of the deleterious selection coefficient ($s$) and population size
($N$), the efficacy of negative selection should depend on demography.
Unfortunately, accommodating demography into theoretic negative selection
models remains an open, difficult problem \parencite{Zeng2013-ep,Johri2020-oj}.
Second, building off classic models of mutation-selection balance
\parencite{Crow1970-wm,Kimura1966-bk}, the BGS model assumes that new mutations
are sufficiently deleterious that they are invariably driven to loss. Under
this assumption, the effect of selection is well-approximated by simply
rescaling the neutral coalescent by a reduction factor known as $B =
\nicefrac{N}{N_e}$ \parencite{Charlesworth2013-kl}. However, this simple
rescaling approach is not appropriate across parts of parameter space that are
relevant to natural populations \parencite{McVean2000-bt,Good2014-yz}. In
particular, the BGS model cannot accommodate the possibility of fixation for
mutations with $2Ns \ll 1$, leading diversity to be incorrectly predicted to
fall to zero as the strength of selection diminishes. This weak selection
problem is related to the wider problem of modeling the substitution rate of
deleterious mutation (i.e. the ``ratchet" rate) and incorporating the impact of
interference from other neighboring selected sites. Finally, the classic BGS
model assumes that selection occurring within other segments does not interfere
with selection occurring in other regions, i.e. no selective or
``Hill--Robertson" interference
\parencite{Hill1966-kd,McVean2000-bt,Felsenstein1974-xm}.

@@

Similar to previous work, we aim to characterize selection processes through
the signal left on nucleotide diversity at linked neutral sites. Although
nucleotide diversity is a less informative statistic than the full site
frequency spectrum, it is much easier to derive theoretic models of how linked
selection alters expected pairwise coalescence times than to describe the
entire genealogy. Still, inferences drawn from the linked selection signal
about selective processes are only as accurate as the underlying theoretic
model. Since we base our method on a fundamentally different theoretic approach
to modeling linked selection, we briefly introduce the theory was extend to
apply across whole-genomes.

Our work extends quantitative genetic models of linked selection
\parencite{Robertson1961-ho,Santiago1995-hx,Santiago1998-bs,Santiago2016-mu},
which model the effects linked selection in terms of polygenic additive fitness
variance ($V_A$). These linked selection models stem from
\textcite{Robertson1961-ho}, which in essence describes how heritable polygenic
fitness variation generates weak genetic draft. At the individual level
Robertson considered, selection generates autocorrelation as offspring from
large families tend to beget many descendents themselves (and likewise with
small families) when fitness is heritable. This same autocorrelation occurs at
the genomic level with linkage \parencite{Santiago1998-bs,Barton2000-zg}, as
the perturbations to a neutral allele's trajectory from its particular fitness
background tend to occur in the same direction across generations until the
background recombines off. These models quantify the total impact of the
autocorrelation generated by weak draft in terms of what we think of as a
\emph{draft-effective} population size $N_d$. The key insight is that in the
long-run, steady, weak draft created by the presence of additive genetic
fitness variance $V_A > 0$ acts like an extra variance in offspring number
would under pure drift \parencite{Wright1938-tv}. However, due to
autocorrelation, the cumulative effect of this fitness variance is inflated by
a factor of $Q^2$. Intuitively, the product $V_A Q^2$ represents the expected
total variance in reproductive success a neutral mutation experiences over its
lifetime in a system with weak selection draft.

Following \textcite{Robertson1961-ho} and \textcite{Santiago1995-hx}, we define
the draft-effective population size $N_d$ by including the total additional
variance created by heritable fitness into Wright's (\citeyear{Wright1938-tv})
equation for effective population size,

@@

Equation \eqref{eq:main_Ne} is general, since different modes of selection and
linkage can be accommodated by different expressions for $V_A$ and the
inflation factor $Q^2$ \parencite{Santiago1995-hx,Santiago1998-bs}. When
fitness variation has a multiplicative polygenic basis, as is often assumed for
genome-wide negative selection processes, the draft asymptotic effective
population size experienced by an arbitrary neutral site under the influence of
all $S$ linked regions is,

@@

The remaining required expressions are the equilibrium additive fitness
variance $V_A$ and the decay rate in associated fitness $1-\kappa$. Fitness
variance could arise from positively or negatively selected alleles, but given
prior work has found negative selection against new deleterious mutations play
a dominant role in shaping genome-wide patterns of diversity
\parencite{McVicker2009-ax,Murphy2022-sj}, we focus on models of fitness
variance created by the continual influx of deleterious mutations into a
population. We imagine a mutation-selection process that creates fitness
variance through a continual influx of deleterious mutations occur at a rate of
$\mu$ per basepair per generation in a conserved region of $L$ basepairs, such
that the region-wide per generation diploid mutation rate is $U = 2 \mu L$.
Each mutation imposes a selective cost of $s$ in heterozygotes and $2s$ in
homozygotes, and fitness effects are multiplicative across sites.

Under this selection model, the additive genic fitness variance created by a
new mutation (at frequency $x=\nicefrac{1}{2N}$) is $2s^2 x(1-x) \approx
\nicefrac{s^2}{N}$. For the entire population of $2N$ chromosomes, the fresh
variance from mutation each generation in segment $i$ is $V_{m,i} \approx
U_is^2$ where $U_i = 2\mu L_i$ is the diploid mutation rate per generation
within the segment. Under the mutation-selection balance assumed by classic BGS
theory, an $L_i$-basepair segment has equilibrium genic variance $V_{a,i}^{BGS}
\approx U_i s$ (see Supplementary Materials Equation \ref{supp-eq:va_bgs}) and
thus $\kappa_i^{BGS} = s$. Substituting $V_{a,i}^{BGS}$ and $\kappa_i^{BGS}$ in
Equation \eqref{eq:polygenic_Ne} and simplifying, we have

@@

However, when new mutations are only weakly deleterious, they can drift up in
frequency before their eventual loss or fixation. At this point, the number of
deleterious mutations is no longer well-approximated by the deterministic
mutation-selection-balance theory; their dynamics are strongly influenced by
stochastic perturbations due to both drift and draft. In this weak selection
regime, classic BGS no longer accurately predicts levels of linked diversity
\parencite{Charlesworth1993-gb,McVean2000-bt,Good2013-lp,Gordo2002-dr}.
Moreover, selection against weakly deleterious mutations alters the
genealogies, such that they are no longer well-approximated by a rescaled
neutral coalescent as under the classic BGS model
\parencite{Przeworski1999-mb,OFallon2010-my,Higgs1995-xc}. To further
complicate matters, the distribution of the number of deleterious mutations
(and its corresponding fitness distribution) is no longer a stationary Poisson
distribution, instead becoming a traveling wave
\parencite{Rouzine2008-qz,Good2013-lp,Gessler1995-hz} towards reduced
population fitness (since we ignore beneficial mutations). When this occurs,
individuals in the least-loaded class are stochastically lost, fixing those
mutations with a click of ``Muller's ratchet"
\parencite{Muller1964-ki,Charlesworth1997-qn}. Unfortunately, determining the
rate of Muller's ratchet is another difficult, open problem
\parencite{Haigh1978-gt,Gordo2002-dr,Gessler1995-hz,Higgs1995-xc} related to
Hill--Robertson interference \parencite{Felsenstein1974-xm}.

Under qualitative genetic models of linked selection, these complications can
can be avoided by finding an expression for the equilibrium fitness variance
that is correct for both strong and weak deleterious mutations.

\textcite{Santiago2016-mu} suggested that this is due to the loss of fitness
variation that occurs when a segregating site fixes and its heterozygosity
becomes zero. If we let $R$ represent the fixation rate (i.e. ratchet rate) in
the region per generation, each fixation removes the equivalent amount of
equilibrium variation put in by mutation. Thus, the steady-state genic variance
under mutation and negative selection is (omitting the segment index $i$ for
clarity),

### 2023-07-21 manuscript/supplementary.tex
ms updates from the frankfurt airport

  \caption{The standard (dark gray) and rescaled (blue) B' maps for different
      $\mu = 1.6 \times 10^{-8}$ and three different selection coefficients.
      For $s=10^{-5}$ ($2N_e s = 0.2$), locally-rescaling alters the predicted
      reduction so that it is essentially insignificant ($B \approx 1$). For
  mid-strength selection $s=10^{-4}$ ($2N_e s = 2$), there is only a very
slight difference between standard and locally-rescaled B' maps. Finally, for
strong selection $s=10^{-3}$ ($2N_e s = 20$), locally-rescaling does not change
the B' maps, as expected. }
\end{figure}

>>>>

  \caption{The standard (dark gray) and locally-rescaled (blue) B' maps for different
      $\mu = 1.6 \times 10^{-8}$ and three different selection coefficients.
      For $s=10^{-5}$ ($2N_e s = 0.2$), locally-rescaling alters the predicted
      reduction so that it is essentially insignificant ($B \approx 1$). For
  mid-strength selection $s=10^{-4}$ ($2N_e s = 2$), there is only a very
slight difference between standard and locally-rescaled B' maps. Finally, for
strong selection $s=10^{-3}$ ($2N_e s = 20$), locally-rescaling does not change
the B' maps, as expected. }
\end{figure}
